ASTR480 Final Report:\\ \textit{Asteroids in TESS}

### 2024-10-20 FinalReport/FinalReportDraft.tex
Almost done

@@

\DeclareRobustCommand{\okina}{%
  \raisebox{\dimexpr\fontcharht\font`A-\height}{%
    \scalebox{0.8}{`}%
  }%
}
\newunicodechar{ʻ}{\okina}
\newcommand{\omuamua}{\okina Oumuamua }

@@

\logo{logos/UC.png}
\school{\textbf{School of Physical and Chemical Sciences}}
\course{\textbf{ASTR480}}
\student{\textbf{Brayden Leicester}}
\supervisor{Dr Michele Bannister and Dr Ryan Ridden-Harper}
\ttitle{Asteroids in TESS} %title of the file

@@

Graduate students, X and Y, helped me understand ... \\

The derivation in section 3 was taken from ... \\

\autoref{Fig:FreqVsDiam} has been recreated after data from the Light Curve Database (LCDB) \citet{Warner2009}, as cited in text. That data has been supplemented with my own work in the corresponding \autoref{Fig:FreqDiamUpdate}.

@@

This is where the abstract goes. \\

This is the next paragraph.

% %*Background

% %*Methods

% %*Main Results

% %*Conclusion

@@

%To get all asteroids in TESS. 
This project aims to find and characterise the light curves of all the asteroids seen by the Transiting Exoplanet Survey Satellite (TESS).
Combining knowledge of these asteroids positions with the high cadence imaging of TESS should allow for the determination of rotation periods of these small bodies, and the amplitude of the variation.

@@

Asteroids are a key class of solar system objects. Much is known about them already, but there is always more to learn.
The orbits of asteroids can be well characterised by only a few values, the proper elements of the object.
These are the lenght of the long axis of elliptical orbit, the semi-major axis, $a$, how elliptical the orbit is, the eccentricity, $e$, and the angle to the ecliptic plane, the inclination,$i$.
Typical values for main belt asteroids are an $a$ of \qtyrange{2}{3.3}{\au}, an $e$ in \qtyrange{0}{0.35}{} and an $i$ around \qtyrange{0}{30}{\degree} \citep{DeMeo2015}.
Comets have undefined $a$ and larger $e$, with some long period comets having $e\approx 1$, but the typical range is \qtyrange{0.2}{0.7}{} \citep{Lewis2012}.
Interstellar objects (ISOs) have $e>1$ as they are on unbound hyperbolic trajectories \citep{McGlynn1989}.

@@

Asteroids can be grouped into families based on clustering of the proper orbital elements and similarities in their colour or spectra \citep{Nesvorny2015}.
These families are named for their largest member.
Families are thought to all originate from the same body that was destroyed by a collision earlier in the solar system's history.

%TODO Rotation of asteroids
As an asteroid rotates, the amount of sunlight reflected changes with time.
This produces a lightcurve that is often sinusoidal in appearance.
These lightcurves are normally ``double-peaked'', having two maximums for one full rotation of the body, due to symmetries in the shape of the object.
The shape of the asteroid has a large effect on the scale of variation in the lightcurve \citep{Durech2015}, the more elongated the object is, the larger the variation.
The rate of rotation is limited by the size of the asteroid \citep{Pravec2000}, as the internal strength of the body is put under too much strain if large asteroid rotate with short periods.

Understanding the rotation properties of asteroids has long been of interest to astronomers \citep[e.g.][for early work into the limits of rotation period and the tumbling nature of some small bodies]{Weidenschilling1981,Harris1994}.
This is still at the forefront of research, with more lightcurves being published year-on-year \citep{Harris2015}.
For the smallest objects ingenious observation techniques are employed to measure rotation periods such as intentionally streaking the asteroid down a column of pixel on a detector, and making the lightcurve from the streak, as done by \citet{Bolin2023}.

Asteroids can have their rotation properties changed with time, collisions can do this, and so can light from the sun.
The YORP effect (named for the discoverers Yarkovsky, O'Keefe, Radzievskii, and Paddack) is the process of tourquing asteroids of an asymmetric shape by both bombardment of photons and thermal (re-)emission \citep{Rubincam2000}.

The data on solar system bodies will only increase with more large survey telescopes coming online soon, such as the Legacy Survey of Space and Time \citep[LSST][]{LSST2019}.
The number of new asteroids that are well characterised will increase by more than 2 orders of magnitude on the Sloan Digital Sky Survey \citep[SDSS][]{York2000} which had $\sim 88000$ objects \citet{Parker2008}.
LSST needs a dedicated pipeline for asteroid classification, the Solar System Notification Alert Processing System \citep[SNAPS][]{Trilling2023} has been developed using Zwicky Transient Facility \citep[ZTF, ][]{Bellm2018} data and can scale up handle volume of data coming from LSST.
Work is being done to understand the biases of how  asymmetric shape high amplitude variations can lead to selection effects of such large sample of asteroids \citet{Levine2023}.

Because of such a rich history of study, many asteroids have a known rotation period.
These are catalogued by \citet{Warner2009} in the Light Curve Database (LCDB).
In the October \nth{1}, 2023 release, the LCDB has nearly 9000 objects with periods determined to a quality code $U\geq 2-$, which is the minimum reliability recommended by \citeauthor{Warner2009} for statistical analysis.
This includes other TESS data from \citet{Pal2020} and \citet{Woods2021}.
The frequencies against diameter plot of these asteroids in \autoref{Fig:FreqVsDiam} shows the collection of all the asteroids in the LCDB with $U\geq 2-$.
The diameter presented in the LCDB here calculated from the absolute magnitude $H$ of the bodies, assuming an albedo.
\citeauthor{Warner2009} acknowledge that there are direct measurements of diameter for more asteroids in the present day, but continue to present the $H$ derived diameters for consistency.

The asteroid spin barrier \citep{Pravec2000} is seen in the red dashed lines in \autoref{Fig:FreqVsDiam}.
There are very few asteroids larger than $\qty{0.15}{\kilo\metre}$ rotating faster than $\sim \qty{10}{\per\day}$.
It is thought that the YORP effect, as well as collisions, spin up asteroids so fast that they pass above this barrier and then no longer have the internal strength necessary to hold themselves together.
This implies that all asteroids made to spin faster have torn themselves apart and that they have no tensile strength, i.e they are rubble piles instead of monoliths \citep{Harris1996}.
This idea of the cohesive strength of small bodies is still explored today \citep[e.g.][for V-type asteroids]{Oszkiewicz2020}.

Comparing the periods I calculate of asteroids that are already in the LCDB is key to making sure my methods are reliable, and that the periods for asteroids for which no data exists can be trusted.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.7\textwidth]{./Figures/Diam-FreqPlot.pdf}
  \caption[Asteroid frequencies vs diameter (LCDB)]{
    Rotation rate of asteroids by their size, recreated from the  October \nth{1}, 2023 LCDB after \citet{Warner2009}.
  }
  \label{Fig:FreqVsDiam}
\end{figure}

\subsubsection*{Interstellar Objects}
%Omuamu
High amplitude variation has come to the forefront of questions about asteroid properties because of the first interstellar object, 1I/\omuamua \citep[see][for a review]{Bannister2019}.
\omuamua was determined to be spectroscopically red \citep{Fitzsimmons2017, Meech2017}, and having a photometric colour in the neutral end of the solar system range \citep{Bannister2017}.
1I was classed as asteroid due to lack of a coma, and no noticeable activity.
This contrasts the second ISO, 2I/Borisov, which had many cometary characteristics \citep[see ][for a review]{Dorofeeva2023}.

\omuamua was measured to have a rotation period of \qty{8.67(0.34)}{\hour} \citep{Belton2018} and seemed to be tumbling \citep[e.g.][]{Drahus2018,Fraser2018}.
Combing the tumbling with an elongation ratio of up to \qty{6(1)}{}:1 \citep{McNeill2018}, 1I is said to have a cigar shape \citep{Belton2018}.
The peak to peak amplitude variation of this ISO was \qty{2.5}{\mag} \citep{Meech2017} over its double-peaked lightcurve, which is interesting as it is more variation than most asteroids more than most asteroids \citep[as seen in the LCDB of][]{Warner2009}.
With the full sky survey of bright asteroids, I hope to and to see just how rare \omuamua is, by finding many asteroids with large amplitude variations.

@@

%Why TESS
TESS is a large area, high cadence imaging, space telescope  \citep{Ricker2014}.
TESS is tasked with observing one piece of sky for \qty{27}{\day} at a time , which defines a sector.
The telescope delivers \qtyproduct{96 x 24}{\degree} full frame images (FFIs) at regular intervals.
These FFIs are built from stacked \qty{2}{\second} exposures, leveraging the short readout times of the 16 charged-coupled devices (CCD) spread over four cameras on-board.
With the initial cadence for these full frame images set to \qty{30}{\minute}, the time resolution of TESS is unparalleled.
As the mission was extended, after TESS had already mapped the entire sky, the length of the FFIs has decreased to \qty{10}{\minute} and then later to \qty{200}{\second}.
This high time resolution and observation area does come at the cost of spatial resolution, as the pixels are each \qty{21}{\arcsec} square.

Of interest here is what such a high sampling rate can do for statistics on the asteroid population.
For bright asteroids the rotation periods should be able to be easily determined from this vast dataset, due to the high sampling of the lightcurve.
The shortest FFIs will be able to accurately determine the rotation periods of all but the fastest rotating asteroids, most of which will be too dim to see in the TESS data.

The highest frequency (shortest period) that can be found from time series data like this is at the Nyquist limit, twice the minimum time between images.
Due to the short cadence, the Nyquist frequency of TESS is quite high, \qty{1}{\per\hour}.
This is well sampled enough to characterise most variable stars, as well as find the orbital periods of exoplanets to a high precision.
The decreased in the FFI exposure time in later sectors comes with a complementary increase in the Nyquist frequency.

Previous work has been conducted to find and classify all the asteroids in TESS data.
\citet{Pal2018} proposes that TESS will be a good instrument for solar system object study, and simulate some detections down to \nth{20} visual magnitude, but that good photometry should only be expected to $V \lesssim 19$, a sentiment shared by \citet{Wong2019}.
The first data release in \citet{Pal2020} catalogues nearly 10,000 asteroid light curves from the first year of TESS operation, and they report that they triple the number of asteroids with accurate rotation periods.
\citet{McNeill2023} also analyse TESS cycle 1, and detect almost 38,000 objects.
They determine reliable rotation periods for about 3,500 asteroids in this sample and show a lack of reliability for objects with periods less than \qty{3}{\hour}.
In the overlap from the data from \citet{Pal2020}, \citeauthor{McNeill2023} find good agreement between the two sets of periods and amplitudes.
The Minor Planet Center (MPC) gets regular updates from TESS thanks to the work of \citet{Woods2021} and their \texttt{LINEAR-TESS} program, which creates tracklets of objects over a day, and then stitches these together to form a track of each asteroid through an FFI.

There is ongoing analysis of subsets of the asteroid population in TESS.
\citet{Gowanlock2024} uses TESS photometry as well as  observations by ZTF to get a longer baseline on mutually observed objects while combining ground and space based observations.
There were not very many objects in both samples with known periods for comparison, with only 222 objects analysed, however they demonstrate that the technique is effective.
Fainter and unknown solar system objects can be found by shift and stacking \citep{Holman2019, Payne2019,Rice2020} or taking a fast X-ray transform \citep{Nguyen2024} of the FFI data.

This work aims to extend these studies to more sectors, hence characterising more asteroids.
The goal is to get a volume complete set of asteroid lightcurves and periods, by analysis of the whole sky as seen by TESS using data from more than just the first year of TESS's operation.
This has not been achieved, as the analysis of only one sector is presented here, but the methods are scalable.
Because of the survey properties, TESS provides a self-consistent way to measure the properties of asteroids over the full sky.
By using the higher cadence FFIs of later TESS sectors, the $P\leq\qty{3}{\hour}$ limit on the accurate periods \citep[as found by ][]{McNeill2023} can be lowered.
I also employ a different data reduction method, using reduced data from the \texttt{TESSreduce} package \citep{Ridden-Harper2021}.
This should increase the accuracy of the photometry due to the improved background subtraction process.

This work is part of a full sky transient survey using TESS, the TESS Extensive Lightcurve Logging and Analysis of Transient Events4 (\texttt{TESSELLATE}) pipeline \citep{TESSELLATE}.
To see what \texttt{TESSELLATE} detects and to help progress the pipeline, one can participate in the Cosmic Cataclysim Zooniverse project\footnote{\href{https://www.zooniverse.org/projects/cheerfuluser/cosmic-cataclysms}{https://www.zooniverse.org/projects/cheerfuluser/cosmic-cataclysms}}.
Asteroids can spike the brightness of a pixel for only a few frames while it is passing by.
This can confuse the pipeline looking for transients such as flare stars and supernova, as they can look very similar, having the brightness change in one pixel for a short time.
Knowing where the asteroids are will allow for the removal of these spikes from the transient pipeline, as well as to understand the asteroid population better.
By filtering the asteroid detections out, while also self consistently determining rotation properties of a large sample of the population, more science can be done with the one set of analyses.

@@

The \texttt{TESSELLATE} pipeline has been running on the OzSTAR supercomputing facilities.
I built this asteroid detection and subsequent lightcurve and periodogram analysis program by testing on one subsection of one CCD of one camera  of one TESS sector.
The code written can be accessed at my ASTR480 GitHub repo \footnote{\href{https://github.com/ble61/ASTR480}{https://github.com/ble61/ASTR480}}.
Because of the large number of asteroids there are too many lightcurves to look at, so the code has to run with no human decisions.
Once I was confident this worked as required, the same code was refactored slightly to work on OzSTAR and a large-scale analysis of a processed TESS sector was run.
% I will be looking for completeness of detections of asteroids, as well as accuracy of periods and amplitude variation.

For this work, I present analysis of TESS sector 29\footnote{\url{https://tess.mit.edu/observations/sector-29/}}, which was observed between \nth{26} August and \nth{22} September 2020.
This was the third TESS sector to be using \qty{10}{\min} cadence FFIs, which has had less analysis of the asteroids preformed.
It was chosen mainly because of this higher cadence, it was the only sector of this cadence that had been processed by \texttt{TESSELLATE} with enough time for my analysis to be preformed.
What is presented for this section is scalable, on a per-sector basis, to any sector that has been reduced.

\subsection{Querrying Databases for Asteroid Positions and Properties}\label{SubSec:Querry}

To check for asteroids in the TESS data, the positions of the asteroids with time are required.
For asteroids bright enough to be in TESS, their ephemeris is well known, so I needed to access their positions and then cross match them with transients in the TESS data.

\texttt{SkyBoT} \citep{Berthier2006} was queried to get positions of asteroids in a cone search.
A box in right ascension (RA) and  declination (Dec) space.
The search was limited to only asteroids with an average visual magnitude $V\geq \qty{20}{\mag}$.
As TESS sectors are \qty{27}{\day} long, so querying every \qty{12}{\hour} is both computationally feasible and allows for a good understanding of a bodies motion.

These positions are still very sparsely spaced in time compared to the TESS data.
An interpolation is needed to bridge the gap, this is preformed by \texttt{numpy} \citep{Harris2020}.

Knowing where the asteroids are is helpful for finding them in the archival TESS data, but knowing more about the individual asteroids is also useful for population statistics.
The \texttt{astroquery} \citep{Ginsburg2019} package was used to query \texttt{JPL Horizons} was used to obtain the orbital elements of each asteroid, $a$, $e$, and $i$, as well as an absolute magnitude $H$, which is a good proxy for size (as discussed above). %TODO ref

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width =0.8 \textwidth]{./Figures/interpPos\_29\_1\_2\_2.pdf}
  \caption[Interpolated Positions of Asteroids]{
    The interpolated positions of asteroids in one subsection of a TESS sector.
    The colours of the lines are time sequenced, as shown in the colour bar.
    The alpha of the colours are scaled by the absolute magnitude $H$ of the asteroid.
    Both celestial coordinates (RA and Dec) and ecliptic longitude and latitude  axes are shown.
    The green $+$ is the centre of the subsection.}
  \label{Fig:interpPos}
\end{figure}

@@

With TESS data coming in $\qty{10}{\minute}$ chunks, 72 interpolated points are needed between each API call.
A linear interpolation between the queried RA and Dec is preformed at the times of the TESS frames.
\texttt{TESSELLATE} saves out these time-frame pairs as part of its analysis, allowing for the time coordinate of any position to be exactly at a frame time.
This will also simplify the detection matching, as the any difference in the time will mean the detection is in a different frame.

These interpolated positions can be seen in \autoref{Fig:interpPos} for a subsection of sector from \texttt{TESSELLATE}.
There are a few interesting features, such as the asteroids are moving in the same direction, indicated by the colouring, they come in from low RA and Dec and tend to increase both coordinates as the month of the sector progresses.
There is also a large size range in this slice of sky, ranging over \qty{5}{\mag} in absolute magnitude, which can be seen in the alpha (or transparency) of the tracks in \autoref{Fig:interpPos}.

>>>>

\usepackage{chngcntr}

@@

\DeclareRobustCommand{\okina}{%
  \raisebox{\dimexpr\fontcharht\font`A-\height}{%
    \scalebox{0.8}{`}%
  }%
}
\newunicodechar{ʻ}{\okina}
\newcommand{\omuamua}{\okina Oumuamua } %space needed to get next word spaced 
\newcommand{\omuamuans}{\okina Oumuamua} %no space after

@@

\logo{logos/UC.png}
\school{\textbf{School of Physical and Chemical Sciences}}
\course{\textbf{ASTR480}}
\student{\textbf{Brayden Leicester}}
\supervisor{Dr Michele Bannister \\ and Dr Ryan Ridden-Harper}
\ttitle{Asteroids in TESS} %title of the file

@@

\autoref{Fig:FreqVsDiam} has been recreated after data from the Light Curve Database (LCDB) \citet{Warner2009}, as cited in text. That data has been supplemented with my own work in the corresponding \autoref{Fig:FreqDiamUpdate}.\\

\autoref{Eq:LCModel} was adapted from the \texttt{Astropy} documentation
\citep{Astropy2022} ( \url{https://docs.astropy.org/en/stable/api/astropy.timeseries.LombScargle.html#astropy.timeseries.LombScargle.model_parameters}) for how the model is fit to the data. \\

The TESS data was reduced for me by \texttt{TESSreduce} as part of the \texttt{TESSELLATE} pipeline, but the photometry is my own.
All the analysis of the photometry was also done by me.

@@

% %*Background
The Transiting Exoplanet Survey Satellite (TESS) is a wealth of high-cadence photometric data. 
Asteroids move across the TESS field, and their intrinsic rotation varies their brightness periodically. 
Here, I present an analysis of the asteroids in TESS sector 29, as an addition to the \texttt{TESSELLATE} pipeline's analysis.

% %*Methods
After obtaining the positions of the asteroids in \qty{12}{\hour} intervals, they are interpolated to match the cadence of the TESS images (\qty{10}{\min}).
Aperture photometry of the asteroids is performed to construct a lightcurve.
The rotation periods and amplitudes are calculated by a Lomb-Scargle periodogram.
Quality checks are applied to properties of both the lightcurve and the periodogram of each object, to ensure accurate quantities are calculated. 
Comparing derived values for objects with known periods, it is found that \qty{70}{\percent} of the detected periods are reliable.

% %*Main Results
5664 asteroids are found in the sector, of which 374 passed all the quality checks.
\qty{80}{\percent} of these have no known period. 
The distribution of periods and amplitudes broadly agrees with other analyses of TESS data.
While some asteroids are calculated to have a large amplitude of variation, manual checking determined this to be an artefact of data contamination.
It is shown that more than half of the objects with a visual magnitude above \qty{16.36(0.73)}{\mag} have reliable periods detected.

% %*Conclusion
While only one sector is analysed here, the methods are shown to be robust and can now be run on any TESS sector.

@@

%To get all asteroids in TESS. 
This project aims to find and characterise the light curves of all the asteroids seen by the Transiting Exoplanet Survey Satellite (TESS).
Combining knowledge of these asteroids positions with the high-cadence imaging of TESS should allow for the determination of rotation periods of these small bodies, and the amplitude of the variation in their lightcurves.

@@

Asteroids are a key class of solar system objects. Much is known about them already, but there is always more to learn.
Three values characterise the orbits of asteroids, the proper elements of the object.
These are the length of the long axis of the orbit, the semi-major axis, $a$, how elliptical the orbit is, the eccentricity, $e$, and the angle to the ecliptic plane, the inclination, $i$.
Typical values for main belt asteroids are an $a$ of \qtyrange{2}{3.3}{\au}, an $e$ in \qtyrange{0}{0.35}{} and an $i$ around \qtyrange{0}{30}{\degree} \citep{DeMeo2015}.
Comets have undefined $a$ and larger $e$, with some long-period comets having $e\approx 1$, but the typical range is \qtyrange{0.2}{0.7}{} \citep{Lewis2012}.
Interstellar objects (ISOs) have $e>1$ as they are on unbound hyperbolic trajectories \citep{McGlynn1989}.

@@

Asteroids can be grouped into families based on  the  clustering of the proper orbital elements and similarities in their colour or spectra \citep{Nesvorny2015}.
These families are named for their largest member.
Families are thought to all originate from the same body that was destroyed by a collision earlier in the solar system's history.

%TODO Rotation of asteroids
As an asteroid rotates, the amount of sunlight reflected changes with time.
This produces a lightcurve that is often sinusoidal in appearance.
These lightcurves are normally ``double-peaked'', with two maximums for one full rotation, due to symmetries in the object's shape. 
The shape of the asteroid has a large effect on the scale of variation in the lightcurve \citep{Durech2015}, the more elongated the object is, the larger the variation.
The rate of rotation is limited by the size of the asteroid \citep{Pravec2000}, as the body's internal strength is placed under too much strain if large
asteroids rotate with short periods.

Understanding the rotation properties of asteroids has long been of interest to astronomers \citep[e.g.][for early work into the limits of rotation period and the tumbling nature of some small bodies]{Weidenschilling1981,Harris1994}.
This is still at the forefront of research, with more lightcurves being published year-on-year \citep{Harris2015}.
For the smallest objects ingenious observation techniques are employed to measure rotation periods such as intentionally streaking the asteroid down a column of pixel on a detector, and making a lightcurve from the streak, as done by \citet{Bolin2023}.

Asteroids can have their rotation properties changed with time, collisions can do this, and so can light from the sun.
The YORP effect (named for the discoverers Yarkovsky, O'Keefe, Radzievskii, and Paddack) is the process of torquing asteroids of an asymmetric shape by both bombardment of photons and thermal (re-)emission \citep{Rubincam2000}.

The long cadence data on solar system bodies will only increase with more large survey telescopes coming online soon, such as the Legacy Survey of Space and Time \citep[LSST][]{LSST2019}.
The number of new asteroids that are well characterised will increase by more than 2 orders of magnitude on the Sloan Digital Sky Survey \citep[SDSS][]{York2000} which had $\sim 88000$ objects \citet{Parker2008}.
LSST needs a dedicated pipeline for asteroid classification, the Solar System Notification Alert Processing System \citep[SNAPS][]{Trilling2023} has been developed using Zwicky Transient Facility \citep[ZTF, ][]{Bellm2018} data and can scale up to handle the volume of data from LSST.
Work is being done to understand the biases of how  asymmetric shape high amplitude variations can lead to selection effects of such a large sample of asteroids \citet{Levine2023}.

Because of such a rich history of study, many asteroids have a known rotation period.
These are catalogued by \citet{Warner2009} in the Light Curve Database (LCDB).
In the October \nth{1}, 2023 release, the LCDB has nearly 9000 objects with periods determined to a quality code $U\geq 2-$, which is the minimum reliability recommended by \citeauthor{Warner2009} for statistical analysis.
This includes other TESS data from \citet{Pal2020} and \citet{Woods2021}.
A frequencies and diameter plot of these asteroids in \autoref{Fig:FreqVsDiam} shows the collection of all the asteroids in the LCDB with $U\geq 2-$.
The diameter presented in the LCDB here is calculated from the absolute magnitude, $H$, of the bodies, assuming an albedo.
\citeauthor{Warner2009} acknowledge that there are direct measurements of diameter for more asteroids in the present day, but continue to present the $H$-derived diameters for consistency.

The asteroid spin barrier \citep{Pravec2000} is seen in the red dashed lines in \autoref{Fig:FreqVsDiam}.
There are very few asteroids larger than $\qty{0.15}{\kilo\metre}$ rotating faster than $\sim \qty{10}{\per\day}$.
It is thought that the YORP effect, as well as collisions, spin up asteroids so fast that they pass above this barrier and then no longer have the internal strength necessary to hold themselves together.
This implies that all asteroids made to spin faster have torn themselves apart because they have no tensile strength, i.e. they are rubble piles instead of monoliths \citep{Harris1996}.
This idea of the cohesive strength of small bodies is still explored today \citep[e.g.][for V-type asteroids]{Oszkiewicz2020}.

Comparing the periods I calculate for asteroids already in the LCDB is an important sanity check of my methods, and ensures the periods calculated for asteroids for which no data exists are reliable.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.7\textwidth]{./Figures/Diam-FreqPlot.pdf}
  \caption[Asteroid frequencies against diameter (LCDB)]{
    Rotation rate of asteroids by their size, recreated from the  October \nth{1}, 2023 LCDB after \citet{Warner2009}.
  }
  \label{Fig:FreqVsDiam}
\end{figure}

\subsubsection*{Interstellar Objects}
High amplitude variation has come to the forefront of questions about asteroid properties because of the first interstellar object, 1I/\omuamua \citep[see][for a review]{Bannister2019}.
\omuamua was determined to be spectroscopically red \citep{Fitzsimmons2017, Meech2017}, and having a photometric colour in the neutral end of the solar system range \citep{Bannister2017}.
1I was classed as an asteroid due to lack of a coma, and no noticeable activity.
This contrasts the second ISO, 2I/Borisov, which had many cometary characteristics \citep[see ][for a review]{Dorofeeva2023}.

\omuamua was measured to have a rotation period of \qty{8.67(0.34)}{\hour} \citep{Belton2018} and seemed to be tumbling \citep[e.g.][]{Drahus2018,Fraser2018}.
Combing the tumbling with an elongation ratio of up to \qty{6(1)}{}:1 \citep{McNeill2018}, 1I is said to have a cigar shape \citep{Belton2018}.
The peak to peak amplitude variation of this ISO was \qty{2.5}{\mag} \citep{Meech2017} over its double-peaked lightcurve, which is interesting as it is more variation than most asteroids more than most asteroids \citep[as seen in the LCDB of][]{Warner2009}.
With the full sky survey of bright asteroids, I aim to see how rare \omuamua is, by finding asteroids with large amplitude variations.

@@

%Why TESS
TESS is a large area, high-cadence imaging, space telescope  \citep{Ricker2014}.
TESS is tasked with observing one piece of sky for \qty{27}{\day} at a time, which defines a sector.
The telescope delivers \qtyproduct{96 x 24}{\degree} full frame images (FFIs).
These FFIs are built from stacked \qty{2}{\second} exposures, leveraging the short readout times of the 16 charged-coupled devices (CCDs) spread over four cameras.
With the initial cadence for these full frame images set to \qty{30}{\minute}, the time resolution of TESS is unparalleled.
As the mission was extended after TESS had already mapped the entire sky, the length of the FFIs decreased to \qty{10}{\minute} and then later to \qty{200}{\second}.
This high time resolution and observation area does come at the cost of spatial resolution, as the pixels are each \qty{21}{\arcsec} square.

Of interest here is what such a high sampling rate can do for statistics on the asteroid population.
Rotation periods of bright asteroids can be easily determined from this vast dataset, due to the high sampling of the lightcurve.
The shortest FFIs will be able to accurately determine the rotation periods of all but the fastest rotating asteroids, most of which will be too dim to see in the TESS data.

The highest frequency (the shortest period) that can be found from time series data is the Nyquist limit, twice the minimum time between images \citep{VanderPlas2018}. 
Due to the short cadence, the Nyquist frequency of TESS is quite high, \qty{1}{\per\hour}.
This is well sampled enough to characterise most variable stars and determine the orbital periods of exoplanets to high precision.
The decrease in the FFI exposure time in later sectors comes with a complementary increase in the Nyquist frequency.

Previous work has been conducted to find and classify all the asteroids in TESS data.
\citet{Pal2018} propose that TESS will be a good instrument for solar system object study, a sentiment shared by \citet{Wong2019}.
\citeauthor{Pal2018} simulate some detections down to \nth{20} visual magnitude and find good photometry should only be expected to $V \lesssim 19$.
The first data release in \citet{Pal2020} catalogues nearly 10,000 asteroid lightcurves from the first year of TESS operation (cycle 1) with the \qty{30}{\min} cadence, and they report that they triple the number of asteroids with accurate rotation periods.
The Minor Planet Center (MPC) gets regular updates from TESS because of the work of \citet{Woods2021} and their \texttt{LINEAR-TESS} program, which creates tracklets of objects over a day, and then stitches these together to form a track of each asteroid through an FFI.

\citet{McNeill2023} also analyse TESS cycle 1, and detect almost 38,000 objects.
They determine reliable rotation periods for about 3,500 asteroids in this sample and show a lack of reliability for objects with periods less than \qty{3}{\hour}.
In the overlap from the data from \citet{Pal2020}, \citeauthor{McNeill2023} find good agreement between the two sets of periods and amplitudes.

There is ongoing analysis of subsets of the asteroid population in TESS.
\citet{Gowanlock2024} use TESS photometry as well as  observations by ZTF to get a longer baseline on mutually observed objects while combining ground and space-based observations.
There were a limited number of objects in both samples with known periods for comparison, with only 222 objects analysed. However, they demonstrate that the technique is effective.
Fainter and unknown solar system objects can be found by shift and stacking \citep{Holman2019, Payne2019,Rice2020} or taking a fast X-ray transform \citep{Nguyen2024} of the FFI data.

This work aims to extend these other TESS studies to more sectors and examine the higher cadence FFIs, hence characterising more asteroids.
The goal is to get a volume complete set of asteroid lightcurves and periods, by analysis of the whole sky as seen by TESS using data from more than just the first year of TESS's operation.
This has not been achieved, as the analysis of only one sector is presented here, but the methods are scalable.

@@

Because of the survey properties, TESS provides a self-consistent way to measure the properties of asteroids over the full sky.
By using the higher cadence FFIs of later TESS sectors, the $P\leq\qty{3}{\hour}$ limit on the accurate periods \citep[as found by ][]{McNeill2023} can be lowered.
I also employ a different data reduction method, using reduced data from the \texttt{TESSreduce} package \citep{Ridden-Harper2021}.
This should increase the accuracy of the photometry due to the improved background subtraction process.

This work is part of a full sky transient survey using TESS, the TESS Extensive Lightcurve Logging and Analysis of Transient Events (\texttt{TESSELLATE}) pipeline \citep{TESSELLATE}.
To see what \texttt{TESSELLATE} detects and to help progress the pipeline, one can participate in the Cosmic Cataclysms Zooniverse project\footnote{\href{https://www.zooniverse.org/projects/cheerfuluser/cosmic-cataclysms}{https://www.zooniverse.org/projects/cheerfuluser/cosmic-cataclysms}}.

Asteroids can spike the brightness of a pixel for only a few frames while it is passing by.
When the pipeline is looking for transients such as flare stars and supernovae, asteroids can look very similar.
There is a brightness change in one pixel for a short time, as the object moves through the field.
Finding the asteroids will allow for these spikes to be filtered from the transient pipeline, while also gaining a better understanding of the asteroid population.
By filtering the asteroid detections out, while also self-consistently determining the rotation properties of a large sample of the population, more science can be done with one set of analyses.

The \texttt{TESSELLATE} pipeline has been running on the OzSTAR supercomputing facilities.
I built this asteroid detection and subsequent lightcurve and periodogram analysis program by testing one subsection of one CCD of one camera  of one TESS sector.
The code written can be accessed at my ASTR480 GitHub repo \footnote{\href{https://github.com/ble61/ASTR480}{https://github.com/ble61/ASTR480}}.
Because of the large number of asteroids there are too many lightcurves to look at, so the code has to run with no human decisions.
Once I was confident this worked as required, the same code was refactored slightly to work on OzSTAR and a large-scale analysis of a processed TESS sector was run.
I have been looking for completeness of detections of asteroids in \texttt{TESSELLATE}, as well as accuracy of periods and amplitude variation calculated.

Here, I present an analysis of TESS sector 29\footnote{\url{https://tess.mit.edu/observations/sector-29/}}, which was observed between \nth{26} August and \nth{22} September 2020.
Sector 29 was the third TESS sector employing \qty{10}{\min} FFIs and this cadence has been focused on less in the literature. 
It was chosen because it was the only sector of this cadence that had been processed by TESSELLATE with enough time for my analysis to be performed. However, what is presented in this section is scalable
on a per-sector basis to any sector that has been reduced.

\subsection{Querying Databases for Asteroid Positions and Properties}\label{SubSec:Querry}

To check for asteroids in the TESS data, the positions of the asteroids with time are required.
For asteroids bright enough to be in TESS, their ephemeris is well known, so I needed to access their positions and then cross-match them with transients in the TESS data.

The sky body tracker, \texttt{SkyBoT} \citep{Berthier2006}, was queried to get the positions of asteroids.
A box in right ascension (RA) and declination (Dec) space was drawn on the sky, centred in the middle of each subsection made by \texttt{TESSELLATE} and a cone search was executed by \texttt{SkyBoT}.
The search was limited to only asteroids with an average visual magnitude $V\geq \qty{20}{\mag}$ because of the analysis of \citet{Pal2018}.
As TESS sectors are \qty{27}{\day} long, querying every \qty{12}{\hour} is computationally feasible and also allows for a good understanding of a body's motion.

Knowing where the asteroids were helped to find them in the archival TESS data, but knowing more about the individual asteroids is also useful for population statistics.
The \texttt{Astroquery} \citep{Ginsburg2019} package was used to query \texttt{JPL Horizons} was used to obtain the orbital elements of each asteroid, $a$, $e$, and $i$, as well as an absolute magnitude $H$.
The proper elements do not change with the asteroid's positions in its orbit, so only one query is needed per object.

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width =0.8 \textwidth]{./Figures/interpPos\_29\_1\_2\_2.pdf}
  \caption[Interpolated Positions of Asteroids on Sky]{
    The interpolated positions of asteroids in one subsection of a TESS sector.
    The colours of the lines are time sequenced, as shown in the colour bar.
    The alpha of the colours are scaled by the absolute magnitude $H$ of the asteroid.
    Both celestial coordinates (RA and Dec) and ecliptic longitude and latitude  axes are shown.
    The green $+$ is the centre of the subsection.}
  \label{Fig:interpPos}
\end{figure}

@@

With TESS data coming in $\qty{10}{\minute}$ chunks, these positions are still very sparsely spaced in time compared to the TESS data.
72 interpolated points are needed between each \texttt{SkyBoT} query.
A linear interpolation between the queried RA and Dec is performed by \texttt{NumPy} \citep{Harris2020} at the times of the TESS frames.
\texttt{TESSELLATE} saves out these time-frame pairs as part of its analysis, allowing for the time coordinate of any position to be exactly at a frame time.
This will also simplify the detection matching, as the any difference in the time will mean the detection is in a different frame.

These interpolated positions can be seen in \autoref{Fig:interpPos} for a subsection of sector from \texttt{TESSELLATE}.
There are a few interesting features, such as the asteroids are moving in the same direction, indicated by the colouring with time.
They tend to decrease their RA and increase their Dec as the month of the sector progresses, generally moving from the bottom right to the top left of the figure.
There is also a large size range in this slice of sky, ranging over \qty{6}{\mag} in absolute magnitude, which can be seen in the alpha (or transparency) of the tracks in \autoref{Fig:interpPos}.

Because of a small change in position between each frame, \citep[$\sim \qty{1}{\px}$ per \qty{30}{min} FFI][]{Pal2018}, this interpolation should be accurate.
Checking this prediction against a higher frequency query to the \texttt{JPL Horizons} ephemeris confirmed this accuracy on a few asteroids, as shown in \autoref{Fig:errPos}.
This is of the asteroid (8654) 1990 KC1, but all other objects checked are just as accurate.
The largest difference is only a third of a TESS pixel in RA, so the interpolations are accurate enough for these measurements.
